The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-06 kalman_walk_paper.tex
edits

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citet{porter2002speciation}
found that speciation could be rapid under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

>>>>

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
Despite the findings of these quantitative genetics models, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found that speciation could be rapid under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
It is known that hybrid incompatibility can result from the increased segregation variance of a quantitative trait \citep{slatkin1994segregation, chevin2014niche}. Here, by focusing on the explicit biological function or physiology of a phenotype, we show that the evolution of even extremely simple phenotypes can appropriately be approximated by quantitative genetics. 
This is because, like the genetic code, the genotype-phenotype ``code'' is fundamentally degenerate -- as we make explicit -- and thus ammendable to quantitative genetic analysis.

Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.
